Impact of asymmetric tail-dependence in extreme climatic events on extinction risk in population ecology

### 2019-04-03 MainText.Rmd
minor edits

<!--Tail dependence of copula and recent result from BIVAN showing LTdep with higher extrisk than UTdep, but 
not explored age structured or in metapopulation context with density dependence effect-->

<!--our objective ?-->
We explored the potential of asymmetric tail-dependence in copula 
for understanding the extinction risk of populations towards extreme climatic event in ecology
by addressing the following specific questions: (Q1) Does nature of tail dependence in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does asymmetric tail-dependent noise across multiple patches influence extinction 
risk of a nonlinear metapopulation model having density dependence? We hypothesized the general finding 
from density-independent metapopulation model [BIVAN paper results of Lewinton-Cohen model]
(higher extinction risk if noise across patches were 
left-tail dependent than the one if noise came from right-tail dependent) holds true only in
the under-compensatory regime, but should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{intro} that how copula structure in noisy 
environment affect the extinction risk of
populations in a single patch age-structured model,
we considered a simple stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation depend upon the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation, respectively.
The environmental stochasticity in the model was embedded in 
the fluctuating vital rates ($f_a$ and $s_e$). We considered the fecundity ($f_a$)
and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. These distribution 
could be generated from iid noises showing specific tail-dependence (left or right tail)
between fecundity and survival rates. Eq. \ref{eq_vitalrates} represent the stochasticity 
in the vital rates. We included the density dependence effect in one of the vital rate,
namely, fecundity ($f_a$). If $\Omega = \infty$ in Eq. \ref{eq_vitalrates}, 
the projection matrix would be independent
of populations from any age-class and the model was then called to be "linear", whereas
its "nonlinear" version had finite $\Omega$.

>>>>

<!--Tail dependence of copula and recent result from BIVAN showing LTdep with higher extrisk than UTdep, but 
not explored in age structured model or in metapopulation context with density dependence effect-->

<!--our objective ?-->
We explored the potential use of asymmetric tail-dependence in copula 
for understanding the extinction risk of populations towards extreme climatic event in ecology
by addressing the following questions: (Q1) Does nature of tail dependence in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does asymmetric tail-dependent noises across multiple patches influence extinction 
risk of a nonlinear metapopulation model with density dependence? Our speculation is that the general finding 
from density-independent metapopulation model [BIVAN paper results of Lewinton-Cohen model]
(higher extinction risk if noises across patches were 
left-tail dependent than the one if noises came with right-tail dependence) holds true only in
the under-compensatory regime, but should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{intro} that how copula structure in noisy 
environment affect the extinction risk of
populations in a single patch age-structured model,
we considered a basic stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation depend upon the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation, respectively.
The environmental stochasticity in the model was embedded in 
the fluctuating vital rates ($f_a$ and $s_e$). We considered the fecundity ($f_a$)
and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. These distributions 
could be generated from iid noises showing specific tail-dependence (left or right tail)
between fecundity and survival rates. Eq. \ref{eq_vitalrates} represent the stochasticity 
in the vital rates. We included the density dependence effect in one of the vital rate,
namely, fecundity ($f_a$). If $\Omega = \infty$ in Eq. \ref{eq_vitalrates}, 
the projection matrix would be independent
of populations from any age-class and the model was then called to be "linear", whereas
its "nonlinear" version had finite $\Omega$.
